Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework

### 2017-05-31 ms/ms.Rnw
Minor edits from James

In this study, we reconcile both intraspecific and interspecific factors and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards. We achieve this by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the capacity of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality rate decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth (i.e. growth-dependent hazard) and death due to stochastic events such as windthrow or fire (i.e. growth-independent or baseline hazard). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}. However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census and growth-dependent and growth-independent hazards?
  \item Does mortality rates vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used 10-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

\begin{figure}[!h]
\includegraphics{figures/fig2.pdf}
\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive accuracy. 
A) Five sequential stages of model selection with each stage increasing in complexity. Null: null model ($\gamma$); Census: inclusion of census effects ($\delta$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{\beta X_{i}})\delta$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{\beta X_{i}} + \gamma)\delta$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Change in logloss under all wood density\textendash parameter combinations. Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. Symbols represent general functional form: null/baseline (circle), growth\textendash dependent hazard model (triangle) and model incorporating both baseline and growth\textendash dependent hazards (square). Note: errors are too small to see in panel A.}
\label{fig2}
\end{figure}

\subsection*{Growth hazard functions}
We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e. $(\alpha e^{\beta X_{i}} + \gamma)\delta$), significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e. $(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig. \ref{fig2}A). Moreover, we found that predictive accuracy was higher for stem diameter growth relative to basal-area growth. Perhaps surprisingly, we detected no significant improvement in predictive accuracy between the baseline only and growth-dependent only hazard functions.

@@

\subsection*{Effect of wood density on mortality rates}
The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig. \ref{fig2} A,B). However, the most parsimonious model with highest predictive accuracy (i.e. lowest logarithmic loss) was that which attributed the wood density effect, $\rho$, to the baseline hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had on higher average mortality rates ((Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species.

@@

\subsection*{Mortality rates, maximum species size and species light requirements}
We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig. \ref{fig6} A,C, E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. The combination of these correlations suggest that species which predominately recruit in gaps (i.e. higher gap index) were more prone to death due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruits in non-gaps. We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig. \ref{fig6} B,D,F,H).

@@

Furthermore, future research should focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}. Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a El-Nino in 1997\textendash 1998. Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf}.

@@

\subsection*{Estimating mortality rates}
At each stage, the best model from the previous stage was taken as input for the next stage. Specifically, stage 1 fitted the simplest model (Fig. \ref{fig1}B) and compared it to the same model but one which varied by census. This allowed us to determine whether $\lambda$ significantly varied over-time and thus whether accounting for it improved predictive capacity. Stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig. \ref{fig1}B) using either annual diameter increment or area growth rates. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure. In stage 3, we begin to examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density. This allowed us to determine if wood density improved predictive capacity, and if so, whether it predominately improved predictive capacity via growth-dependent or growth-independent processes. Stage 4 also examined inter-specific variability, however, here the focus was on using the best model of stage 2 and estimating the total predictive gain obtained if all parameters were allowed to vary as a function of species ID. The last stage was fitting the best model, with species random effects and wood density to the entire dataset (i.e. not cross validated). This allowed us to examine the range of species hazard curves, how they are affected by wood density, census and posthoc covariates, and the proportion of predicted variation explained by species, wood density, census and growth-dependent and baseline hazard terms.

>>>>

In this study, we reconcile both intraspecific and interspecific factors and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards. We achieve this by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the capacity of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality rate decaying exponentially towards an asymptote at zero. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth (i.e. growth-dependent hazard) and death due to stochastic events such as windthrow or fire (i.e. growth-independent or baseline hazard). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}. However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census and growth-dependent and growth-independent hazards?
  \item Does mortality rates vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used 10-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increased model complexity (see methods).

@@

\begin{figure}[!h]
\includegraphics{figures/fig2.pdf}
\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive accuracy. 
A) Five sequential stages of model selection with each stage increasing in complexity. Null: null model ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate. This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
B) Change in logloss under all wood density\textendash parameter combinations. Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth. Symbols represent general functional form: null/baseline (circle), growth\textendash dependent hazard model (triangle) and model incorporating both baseline and growth\textendash dependent hazards (square). Note: errors are too small to see in panel A.}
\label{fig2}
\end{figure}

\subsection*{Growth hazard functions}
We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), significantly outperformed both the baseline only (i.e.$\gamma \delta_t$) or growth-dependent only (i.e. $(\alpha e^{-\beta X_{i}})\delta_t$) hazard functions (Fig. \ref{fig2}A). Moreover, we found that predictive accuracy was higher for stem diameter growth relative to basal-area growth. Perhaps surprisingly, we detected no significant improvement in predictive accuracy when examining models that only accounted for baseline or growth-dependent only hazards.

@@

\subsection*{Effect of wood density on mortality rates}
The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig. \ref{fig2} A,B). However, the most parsimonious model with highest predictive accuracy (i.e. lowest logarithmic loss) was that which attributed the wood density effect, $\rho$, to the baseline hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had on higher average mortality rates (Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species.

@@

\subsection*{Mortality rates, maximum species size and species light requirements}
We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were weakly and positively correlated with gap index (our measure of species light requirement; Fig. \ref{fig6} A,C, E). By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index. The combination of these correlations suggest that species which predominately recruit in gaps (i.e. higher gap index) were more prone to death due to both low growth and stochastic chance. They also required faster growth rates to achieve a given mortality rate relative to a species that recruits in non-gaps. We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig. \ref{fig6} B,D,F,H).

@@

Furthermore, future research should focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}. Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010. Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3). However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by an El-Nino in 1997\textendash 1998. Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf}.

@@

\subsection*{Estimating mortality rates}
At each stage, the best model from the previous stage was taken as input for the next stage. Specifically, stage 1 fitted the simplest model (Fig. \ref{fig1}B) and compared it to the same model but one which varied by census. This allowed us to determine whether $\lambda$ significantly varied over-time and thus whether accounting for it improved predictive capacity. Stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig. \ref{fig1}B) using either annual diameter increment or area growth rates. This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure. In stage 3, we examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density. This allowed us to determine if wood density improved predictive capacity, and if so, whether it predominately improved predictive capacity via growth-dependent or growth-independent processes. Stage 4 also examined inter-specific variability, however, here the focus was on using the best model of stage 2 and estimating the total predictive gain obtained if all parameters were allowed to vary as a function of species ID. The last stage was fitting the best model, with species random effects and wood density to the entire dataset (i.e. not cross validated). This allowed us to examine the range of species hazard curves, how they are affected by wood density, census and posthoc covariates, and the proportion of predicted variation explained by species, wood density, census and growth-dependent and baseline hazard terms.
